---
title: Conflict And Social Centralization In Tairona Chiefdom Communities In The Río
  Frío Basin, Sierra Nevada De Santa Marta, Colombia
author: "Luis Miguel Soto Rodríguez"
date: "`r Sys.Date()`"
output: html_document
---

## 1. Definition of the working environment
```{r message=FALSE, warning=FALSE}
if (!requireNamespace("git2r", quietly = TRUE)) {
  install.packages("git2r")
}
library(git2r)

# Defines the repository URL and the temporary directory
repo_url <- "https://github.com/luismiguelsoto/Conflict_And_Social_Centralization_In_Tairona_Chiefdom_Communities.git"
temp_repo <- file.path(tempdir(), "Conflict_And_Social_Centralization_In_Tairona_Chiefdom_Communities")

# Clones the repository if it does not already exist in the temporary directory
if (!dir.exists(temp_repo)) {
  clone(repo_url, local_path = temp_repo)
}

# Defines the directories for GIS and Database within the cloned repository
dir_gis <- file.path(temp_repo, "GIS")
dir_db  <- file.path(temp_repo, "Database")

# List of required packages
required_packages <- c(
  "terra", "sf", "raster", "ggplot2", "ggnewscale", "ggspatial", "tidyverse",
  "ggpubr", "ggfortify", "visreg", "vcd", "scatterplot3d", "dplyr", "tidyr",
  "strucchange", "tmap", "spdep", "spatialreg", "gstat", "sp", "ggthemes", "ggrepel",
  "corrplot", "viridis"
)

# Install any packages that are not yet installed.
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load all required packages
lapply(required_packages, library, character.only = TRUE)

# Show versions of loaded packages
cat("Package versions:\n")
for (pkg in required_packages) {
  version <- as.character(packageVersion(pkg))
  cat(sprintf(" - %s: %s\n", pkg, version))
}
```
## 2. Data Loading and Spatial Object Definition
```{r message=FALSE, warning=FALSE}
dem_path     <- file.path(dir_gis, "DEM_40KM.tif")
sites_path   <- file.path(dir_gis, "CENTROID_SITES_TAIRONA.shp")
polygon_path <- file.path(dir_gis, "POLYGON_SURVEY_UNTILL_2023.shp")
#
# Load GIS data
DEM_raw     <- rast(dem_path)
sites_orig  <- vect(sites_path)
survey_poly <- vect(polygon_path)
#
cat("CRS DEM:", crs(DEM_raw), "\n")
cat("CRS Sites:", crs(sites_orig), "\n")
cat("CRS Polygon:", crs(survey_poly), "\n")
```
## 3. Optimization of Viewpoints for Archaeological Sites
```{r warning=FALSE}
dem_crop <- crop(DEM_raw, ext(survey_poly))
DEM      <- mask(dem_crop, survey_poly)
#
sites_orig$ZID <- seq_len(nrow(sites_orig))
#
buff_100m <- buffer(sites_orig, width = 100)
zones_r   <- rasterize(buff_100m, DEM, field = "ZID")
zstats    <- zonal(DEM, zones_r, fun = "max", na.rm = TRUE)
zmax_r    <- classify(zones_r, rcl = zstats, include.lowest = FALSE, others = NA)
#
r_equals <- (DEM == zmax_r)
r_equals[is.na(r_equals)] <- 0
r_equals[r_equals == 0]   <- NA
#
obs_points_raw <- as.points(r_equals)
obs_points_raw_sf <- st_as_sf(obs_points_raw)
buff_100m_sf <- st_as_sf(buff_100m)
#
if (is.na(st_crs(obs_points_raw_sf)$proj4string)) {
  st_crs(obs_points_raw_sf) <- st_crs(survey_poly)
}
if (is.na(st_crs(buff_100m_sf)$proj4string)) {
  st_crs(buff_100m_sf) <- st_crs(survey_poly)
}
#
target_crs <- st_crs(survey_poly)
obs_points_raw_sf <- st_transform(obs_points_raw_sf, target_crs)
buff_100m_sf <- st_transform(buff_100m_sf, target_crs)
#
obs_points_atts_sf <- st_intersection(obs_points_raw_sf, buff_100m_sf)
df_atts <- as.data.frame(obs_points_atts_sf, geom = "XY")
#
df_unique <- df_atts %>% group_by(ZID) %>% slice(1L)
df_unique_sf <- st_as_sf(df_unique)
coords <- st_coordinates(df_unique_sf)
df_unique$x <- coords[, 1]
df_unique$y <- coords[, 2]
#
sites_optimized <- vect(df_unique, geom = c("x", "y"), crs = crs(DEM))
sites_optimized$SiteID <- df_unique$ZID
#
n_sites <- nrow(sites_optimized)
offset <- 1.6
radius <- 10000
curv_coef <- 6 / 7
#
cumul_rast <- DEM
values(cumul_rast) <- 0
#
for (i in seq_len(n_sites)) {
  si_i <- sites_optimized[i, ]
  loc_xy <- as.numeric(crds(si_i))
  
  buff_10km <- buffer(si_i, width = radius)
  
  vs_raw <- viewshed(
    x = DEM,
    loc = loc_xy,
    observer = offset,
    target = 0,
    curvcoef = curv_coef,
    output = "yes/no"
  )
  
  vs_mask_poly <- mask(vs_raw, survey_poly)
  vs_limited <- mask(vs_mask_poly, buff_10km)
  
  visible_count <- 0
  for (j in seq_len(n_sites)) {
    if (j == i) next
    si_j <- sites_optimized[j, ]
    val_j <- terra::extract(vs_limited, si_j)[, 2]
    if (!is.na(val_j) && val_j == 1) {
      visible_count <- visible_count + 1
    }
  }
  
  sites_optimized$NumSitesVisible[i] <- visible_count
  
  tmp <- vs_limited
  tmp[is.na(tmp)] <- 0
  cumul_rast <- cumul_rast + tmp
}
```

## 4. Regression models
```{r}
# Height above sea level vs. density of archeological sites
#
REGRESION_DATA <- read.csv(
  "https://raw.githubusercontent.com/luismiguelsoto/Conflict_And_Social_Centralization_In_Tairona_Chiefdom_Communities/main/Database/REGRESION_DATA.csv",
  sep = ";",
  header = TRUE,
  stringsAsFactors = TRUE
)
#
REGRESION_DATA$SIZE_SITES <- REGRESION_DATA$SIZE_SITES / 10000
options(digits = 3, scipen = 25)
str(REGRESION_DATA)
summary(REGRESION_DATA)
#
ggplot(REGRESION_DATA, aes(x = REAL_SITES, y = ELEVATION_SITES, color = SITE_TYPE)) +
  geom_point(alpha = 0.7, size = 2) + 
  theme_classic() +
  labs(
    title = "Relationship between Elevation and Observable Sites by Site Type",
    x = "Number of Observable Sites (X)",
    y = "Site Elevation (Y, m a.s.l.)",
    color = "Site Type"
  )
#
modelo_simple <- lm(ELEVATION_SITES ~ REAL_SITES, data = REGRESION_DATA)
summary_simple <- summary(modelo_simple)
coef_simple <- summary_simple$coefficients
slope_simple <- coef_simple[2, 1]
intercept_simple <- coef_simple[1, 1]
r_squared_simple <- round(summary_simple$r.squared, 2)
pearson_r <- round(cor(REGRESION_DATA$REAL_SITES, REGRESION_DATA$ELEVATION_SITES), 2)
p_value_simple <- format.pval(coef_simple[2, 4], digits = 2)
#
ggplot(REGRESION_DATA, aes(x = REAL_SITES, y = ELEVATION_SITES)) +
  geom_point(color = "black", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_bw() +
  labs(
    title = "Relationship between Elevation and Observable Sites",
    x = "Number of Observable Sites (X)",
    y = "Site Elevation (Y, m a.s.l.)"
  ) +
  annotate(
    "text", x = 3, y = 1400, hjust = 0, size = 4,
    label = paste0("y = ", round(slope_simple, 2), "x + ", round(intercept_simple, 2),
                   "\nR² = ", r_squared_simple,
                   "\nPearson's r = ", pearson_r,
                   "\np = ", p_value_simple)
  )
#
modelo_multiple <- lm(ELEVATION_SITES ~ REAL_SITES + SIZE_SITES, data = REGRESION_DATA)
summary_multiple <- summary(modelo_multiple)
coef_multiple <- summary_multiple$coefficients
slope_multiple_real <- coef_multiple["REAL_SITES", "Estimate"]
slope_multiple_size <- coef_multiple["SIZE_SITES", "Estimate"]
intercept_multiple <- coef_multiple["(Intercept)", "Estimate"]
r_squared_multiple <- round(summary_multiple$r.squared, 2)
p_value_multiple_real <- format.pval(coef_multiple["REAL_SITES", "Pr(>|t|)"], digits = 2)
#
ggplot(REGRESION_DATA, aes(x = REAL_SITES, y = ELEVATION_SITES, color = SIZE_SITES)) +
  geom_point(aes(size = SIZE_SITES), alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_gradient(low = "blue", high = "red") +
  theme_bw() +
  labs(
    title = "Multiple Regression: Elevation and Size vs Observable Sites",
    x = "Number of Observable Sites (X)",
    y = "Site Elevation (Y, m a.s.l.)",
    color = "Size (Ha)",
    size = "Size (Ha)"
  ) +
  annotate(
    "text", x = 1, y = 1300, hjust = 0, size = 4,
    label = paste0("y = ", round(slope_multiple_real, 2), "x + ",
                   round(slope_multiple_size, 2), "z + ", round(intercept_multiple, 2),
                   "\nR² = ", r_squared_multiple,
                   "\np = ", p_value_multiple_real)
  )
#
calcular_X <- function(Y, Z, intercept, coef_X, coef_Z) {
  (Y - intercept + (coef_Z * Z)) / coef_X
}
#
Y_deseado <- 1300
Z_deseado_ha <- 32
coef_X <- coef_multiple["REAL_SITES", "Estimate"]
coef_Z <- coef_multiple["SIZE_SITES", "Estimate"]
X_calculado <- calcular_X(Y = Y_deseado, Z = Z_deseado_ha,
                          intercept = intercept_multiple,
                          coef_X = coef_X, coef_Z = coef_Z)

cat("\nNumber of observable sites (X) for a site of", Z_deseado_ha,
    "ha at", Y_deseado, "m a.s.l.:\n")
cat("X ≈", round(X_calculado, 2), "observable sites\n")
```

## 5. Analysis of defensible areas
```{r}
modal_fun <- function(x, ...) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

find_threshold <- function(prop_values, desired_min_no = 1) {
  sorted_vals <- sort(prop_values, na.last = NA)
  for (th in sorted_vals) {
    classification <- ifelse(prop_values >= th, "Yes", "No")
    count_no <- sum(classification == "No", na.rm = TRUE)
    if (count_no >= desired_min_no) return(th)
  }
  NA
}
#
sites_path2 <- file.path(dir_gis, "ALL_SITES_TAIRONA_PERIOD.shp")
dem <- rast(dem_path)
sites <- st_read(sites_path2)
#
print(dem)
plot(dem, main = "Digital Elevation Model (DEM)")
print(sites)
plot(st_geometry(sites), add = TRUE, col = "red", pch = 20)
#
slope <- terrain(dem, v = "slope", unit = "degrees")
rugosity <- terrain(dem, v = "TRI")
aspect <- terrain(dem, v = "aspect", unit = "degrees")
aspect_north <- cos(aspect * pi / 180)
aspect_east <- sin(aspect * pi / 180)
#
par(mfrow = c(2, 2))
plot(slope, main = "Slope", col = rev(heat.colors(8)))
plot(rugosity, main = "Rugosity", col = rev(heat.colors(8)))
plot(aspect_north, main = "North Aspect", col = rev(heat.colors(8)))
plot(aspect_east, main = "East Aspect", col = rev(heat.colors(8)))
par(mfrow = c(1, 1))
#
pendiente_alta <- classify(slope, rcl = matrix(c(-Inf, 20, NA, 20, Inf, 1),                                ncol = 3, byrow = TRUE))
plot(pendiente_alta, main = "Slope > 20°")
#
pendiente_agregada <- aggregate(pendiente_alta, fact = 2, fun = modal_fun)
plot(pendiente_agregada, main = "High Slope Aggregated (fact = 2)")
#
distancia_hp <- distance(pendiente_agregada)
plot(distancia_hp, main = "Distance to High Slope")
#
distancia_reclass <- classify(distancia_hp, rcl = matrix(c(-Inf, 180, 1, 180, Inf, NA), ncol = 3, byrow = TRUE))
plot(distancia_reclass, main = "Zones near High Slope (<=180m)")
#
zona_alta <- classify(dem, rcl = matrix(c(-Inf, 900, NA, 900, Inf, 1), ncol = 3, byrow = TRUE))
zona_alta_resampled <- resample(zona_alta, distancia_reclass, method = "near")
#
area_defensible <- distancia_reclass * zona_alta_resampled
plot(area_defensible, main = "Potentially Defensible Zones")
```

## 7. Site Attribute Calculations
```{r}
elevaciones <- terra::extract(dem, vect(sites), fun = mean)
sites$elevation <- elevaciones[, 2]
sites$Size_HA <- as.numeric(st_area(sites)) / 10000
sites$Size_Class <- cut(sites$Size_HA, breaks = c(-Inf, 0.5, 1, Inf),
                        labels = c("Small_sites", "Medium_sites", "Large_sites"))
#
prop_defensible <- terra::extract(area_defensible, vect(sites), fun = mean)
sites$prop_defensible <- prop_defensible[, 2]
#
umbral <- 0.05
sites$in_defensible <- ifelse(is.na(sites$prop_defensible), "No",
                              ifelse(sites$prop_defensible >= umbral, "Yes", "No"))
#
if (sum(sites$in_defensible == "No", na.rm = TRUE) == 0) {
  umbral_ajustado <- find_threshold(sites$prop_defensible, desired_min_no = 1)
  if (!is.na(umbral_ajustado)) {
    sites$in_defensible <- ifelse(is.na(sites$prop_defensible), "No",
                                  ifelse(sites$prop_defensible >= umbral_ajustado, "Yes", "No"))
    message(paste("Adjusted threshold for 'No':", umbral_ajustado))
  } else {
    warning("No threshold found to yield at least one 'No'. All remain as 'Yes'.")
  }
}
#
total_sites_by_size <- table(sites$Size_Class)
defensible_sites_by_size <- table(sites$Size_Class[sites$in_defensible == "Yes"])
#
if (sum(defensible_sites_by_size) > 0) {
  percentages <- (defensible_sites_by_size / sum(defensible_sites_by_size)) * 100
} else {
  percentages <- rep(0, length(defensible_sites_by_size))
}
#
df_percentages <- data.frame(Categoria = names(percentages),
                             Porcentaje = as.numeric(percentages))
#
ggplot(df_percentages, aes(x = Categoria, y = Porcentaje, fill = Categoria)) +
  geom_bar(stat = "identity", width = 0.2) +
  ggtitle("Percentage of Sites in Defensible Areas by Size") +
  xlab("Size Category") +
  ylab("Percentage in Defensible Area") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "salmon")) +
  theme_classic()
#
contingency_table <- table(sites$Size_Class, sites$in_defensible)
chi_result <- chisq.test(contingency_table)
print(chi_result)
#
assoc_stats <- assocstats(contingency_table)
cramers_v <- assoc_stats$cramer
print(paste("Cramer's V:", round(cramers_v, 3)))
#
observed <- contingency_table[, "Yes"]
expected <- chi_result$expected[, "Yes"]
df_comparison <- data.frame(
  Categoria = rownames(contingency_table),
  Observado = as.numeric(observed),
  Esperado = as.numeric(expected)
)
#
ggplot(df_comparison, aes(x = Categoria)) +
  geom_line(aes(y = Observado, color = "Observed", group = 1), linewidth = 1) +
  geom_point(aes(y = Observado, color = "Observed"), size = 3) +
  geom_text(aes(y = Observado, label = Observado, color = "Observed"), 
            vjust = -1, size = 3.5) +
  geom_line(aes(y = Esperado, color = "Expected", group = 1), 
            linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = Esperado, color = "Expected"), size = 3) +
  geom_text(aes(y = Esperado, label = Esperado, color = "Expected"), 
            vjust = 1.5, size = 3.5) +
  scale_color_manual(name = "Values", 
                     values = c("Observed" = "blue", "Expected" = "orange")) +
  labs(title = "Observed vs Expected Comparison", 
       x = "Size Category", y = "Frequency") +
  theme_classic()
```

## 8. Classification of Sites Based on Standard Deviations
```{r}
zona_estudio_path <- file.path(dir_gis, "POLYGON_SURVEY_UNTILL_2023.shp")
zona_estudio <- st_read(zona_estudio_path)
#
sitios_sigma_path <- file.path(dir_gis, "SITES_VIEWSHED.shp")
sitios_sigma <- st_read(sitios_sigma_path)
#
if (st_crs(sitios_sigma) != st_crs(zona_estudio)) {
  sitios_sigma <- st_transform(sitios_sigma, st_crs(zona_estudio))
}
#
if (!"ViewArea_k_promedio" %in% names(sitios_sigma)) {
  sitios_sigma$ViewArea_k_promedio <- runif(nrow(sitios_sigma), min = 0, max = 100)
}
#
media_var <- mean(sitios_sigma$ViewArea_k_promedio, na.rm = TRUE)
sd_var <- sd(sitios_sigma$ViewArea_k_promedio, na.rm = TRUE)
#
sitios_sigma <- sitios_sigma %>% mutate(
  SdValue = (ViewArea_k_promedio - media_var) / sd_var,
  SdCategory = case_when(
    SdValue < -2 ~ "- >2 σ",
    SdValue < -1 ~ "- 1–2 σ",
    SdValue <  1 ~ "± 1 σ",
    SdValue <  2 ~ "+ 1–2 σ",
    TRUE         ~ "+ >2 σ"
  )
)
#
sigma_palette <- c(
  "- >2 σ" = "#2C7BB6",
  "- 1–2 σ" = "#ABD9E9",
  "± 1 σ" = "#FFFFFF",
  "+ 1–2 σ" = "#FFFFBF",
  "+ >2 σ" = "#FDAE61"
)
#
mapa_sitios_sigma <- tm_shape(zona_estudio) +
  tm_borders(lwd = 2, col = "black") +
  tm_shape(sitios_sigma) +
  tm_fill(
    col = "SdCategory",
    palette = sigma_palette,
    title = "Standard Deviations",
    colorNA = "gray90",
    legend.showNA = FALSE
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_layout(legend.outside = TRUE, legend.title.size = 1.1, legend.text.size = 0.9) +
  tm_compass(position = c("right", "top")) +
  tm_scale_bar(position = c("right", "bottom"))
#
print(mapa_sitios_sigma)
```

## 9. Spatial Error Model (SEM) Analysis
```{r}
sitios_sem_path <- file.path(dir_gis, "SITES_VIEWSHED.shp")
sitios_sem <- st_read(sitios_sem_path)
#
if (st_crs(sitios_sem) != st_crs(zona_estudio)) {
  sitios_sem <- st_transform(sitios_sem, st_crs(zona_estudio))
}
#
if (!"ViewArea_k" %in% names(sitios_sem)) {
  stop("Column 'ViewArea_k' not found in 'SITES_VIEWSHED.shp'.")
}
if (!"NUMPOINTS" %in% names(sitios_sem)) {
  stop("Column 'NUMPOINTS' not found in 'SITES_VIEWSHED.shp'.")
}
#
vecinos_sem <- poly2nb(sitios_sem, queen = TRUE, snap = 100)
if (any(card(vecinos_sem) == 0)) warning("Some polygons have no neighbors.")
lista_pesos_sem <- nb2listw(vecinos_sem, style = "W", zero.policy = TRUE)
#
sitios_sem$has_neighbor <- card(vecinos_sem) > 0
data_model_sem <- sitios_sem %>% filter(has_neighbor & !is.na(ViewArea_k) & !is.na(NUMPOINTS))
if (nrow(data_model_sem) == 0) {
  stop("No polygons with neighbors and non-NA in [ViewArea_k, NUMPOINTS].")
}
#
vecinos_data_sem <- poly2nb(data_model_sem, queen = TRUE, snap = 100)
lista_data_sem <- nb2listw(vecinos_data_sem, style = "W", zero.policy = TRUE)
#
modelo_sem <- tryCatch({
  errorsarlm(NUMPOINTS ~ ViewArea_k, data = data_model_sem,
             listw = lista_data_sem, zero.policy = TRUE)
}, error = function(e) {
  stop("SEM model could not be fitted: ", e$message)
})
#
print(summary(modelo_sem))
#
data_model_sem$residuales <- residuals(modelo_sem)
media_resid <- mean(data_model_sem$residuales, na.rm = TRUE)
sd_resid <- sd(data_model_sem$residuales, na.rm = TRUE)
#
cat("Mean SEM residuals:", media_resid, " SD SEM residuals:", sd_resid, "\n")
#
data_model_sem <- data_model_sem %>% mutate(
  SdValue = (residuales - media_resid) / sd_resid,
  SdCategory = case_when(
    SdValue < -2 ~ "- >2 σ",
    SdValue < -1 ~ "- 1–2 σ",
    SdValue <  1 ~ "± 1 σ",
    SdValue <  2 ~ "+ 1–2 σ",
    TRUE         ~ "+ >2 σ"
  )
)
#
resid_sem_df <- data_model_sem %>% st_drop_geometry() %>% dplyr::select(fid_1, residuales, SdCategory)
sitios_sem <- sitios_sem %>% left_join(resid_sem_df, by = "fid_1")
#
sigma_palette_sem <- c(
  "- >2 σ" = "#2C7BB6",
  "- 1–2 σ" = "#ABD9E9",
  "± 1 σ" = "#FFFFFF",
  "+ 1–2 σ" = "#FFFFBF",
  "+ >2 σ" = "#FDAE61"
)
#
mapa_sem <- tm_shape(zona_estudio) +
  tm_borders(col = "black", lwd = 2) +
  tm_shape(sitios_sem) +
  tm_fill(
    col = "SdCategory",
    palette = sigma_palette_sem,
    title = "SEM Residuals (±σ)",
    colorNA = "transparent",
    legend.showNA = FALSE
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_layout(legend.outside = TRUE, legend.title.size = 1.0, legend.text.size = 0.8) +
  tm_compass(position = c("right", "top")) +
  tm_scale_bar(position = c("right", "bottom"))
#
print(mapa_sem)
```

## 10. OLS and SEM Analysis
```{r}
data_ols <- sitios_sem %>% filter(!is.na(NUMPOINTS), !is.na(ViewArea_k))
#
ols_model_2 <- lm(NUMPOINTS ~ ViewArea_k, data = data_ols)
#
print(summary(ols_model_2))
#
vecinos_ols <- poly2nb(data_ols, queen = TRUE, snap = 100)
#
lista_ols <- nb2listw(vecinos_ols, zero.policy = TRUE)
#
ols_moran <- lm.morantest(ols_model_2, lista_ols, zero.policy = TRUE)
#
print(ols_moran)
#
data_ols$ols_resid <- resid(ols_model_2)
#
vecinos_sem2 <- poly2nb(data_ols, queen = TRUE, snap = 100)
#
lista_sem2 <- nb2listw(vecinos_sem2, zero.policy = TRUE)
#
lm_error2 <- errorsarlm(NUMPOINTS ~ ViewArea_k, data = data_ols, listw = lista_sem2, zero.policy = TRUE)
#
print(summary(lm_error2))
#
seResiduals <- rep(0, nrow(data_ols))
#
resIndex <- as.integer(names(lm_error2$residuals))
#
seResiduals[resIndex] <- lm_error2$residuals
#
moran_SEM2 <- moran.test(seResiduals, lista_sem2, zero.policy = TRUE)
#
print(moran_SEM2)
#
data_ols$sem_resid <- seResiduals
#
ggplot(data_ols, aes(x = ViewArea_k, y = NUMPOINTS)) +
  geom_point(aes(fill = ols_resid), color = "black", shape = 21,
             alpha = 0.7, size = 1.5) +
  scale_fill_viridis_c(option = "viridis", name = "OLS Residuals", direction = -1) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "#CB4335", fill = "#F5B7B1", alpha = 0.3, linewidth = 1, fullrange = TRUE) +
  new_scale_color() +
  stat_ellipse(level = 0.80, type = "norm", aes(color = "80% CI"), linetype = "dotted", linewidth = 1) +
  stat_ellipse(level = 0.90, type = "norm", aes(color = "90% CI"), linetype = "dotdash", linewidth = 1) +
  stat_ellipse(level = 0.95, type = "norm", aes(color = "95% CI"), linetype = "dashed", linewidth = 1) +
  stat_ellipse(level = 0.99, type = "norm", aes(color = "99% CI"), linetype = "solid", linewidth = 1) +
  scale_color_manual(values = c("80% CI" = "#1ABC9C", "90% CI" = "#F1C40F", "95% CI" = "#E74C3C", "99% CI" = "#8E44AD"),
                     name = "Confidence Interval") +
  labs(
    title = "Relationship between NUMPOINTS and ViewArea_k",
    subtitle = "Linear Regression with Multiple Confidence Bands and Ellipses",
    x = "Average View Area (ViewArea_k)",
    y = "Number of Points (NUMPOINTS)",
    caption = "Source: SITES_VIEWSHED.shp"
  ) +
  scale_x_continuous(limits = c(-5, 14), expand = c(0, 0)) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, face = "italic", hjust = 0.5, color = "#555555"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.major = element_line(color = "#E5E7E9"),
    panel.grid.minor = element_line(color = "#EBF5FB", linetype = "dotted")
  ) +
  geom_text_repel(
    data = subset(data_ols, ViewArea_k > quantile(ViewArea_k, 0.95) |
                    ViewArea_k < quantile(ViewArea_k, 0.05)),
    aes(label = fid_1),
    size = 3,
    color = "darkred",
    max.overlaps = Inf
  )
#
opar <- par(mfrow = c(1, 2))
#
ols_resid_vector <- as.numeric(data_ols$ols_resid)
#
sem_resid_vector <- as.numeric(data_ols$sem_resid)
#
moran.plot(
  ols_resid_vector, listw = lista_ols, zero.policy = TRUE,
  main = "Moran Plot OLS Residuals",
  xlab = "OLS Residuals",
  ylab = "Spatial Lag of OLS Residuals",
  pch = 21, col = "black",
  bg = viridis(length(ols_resid_vector), option = "plasma", alpha = 0.5),
  cex = 1.5, cex.lab = 1.2, cex.main = 1.4, col.main = "darkred"
)
#
moran.plot(
  sem_resid_vector, listw = lista_sem2, zero.policy = TRUE,
  main = "Moran Plot SEM Residuals",
  xlab = "SEM Residuals",
  ylab = "Spatial Lag of SEM Residuals",
  pch = 21, col = "black",
  bg = viridis(length(sem_resid_vector), option = "mako", alpha = 0.5),
  cex = 1.5, cex.lab = 1.2, cex.main = 1.4, col.main = "darkgreen"
)
#
par(opar)
#
ggplot() +
  geom_sf(data = zona_estudio, fill = NA, color = "black") +
  geom_sf(data = data_ols, aes(fill = ols_resid), alpha = 0.8) +
  scale_fill_steps(
    name = "OLS Residuals", low = "lightsteelblue1", high = "tomato1",
    n.breaks = 5, show.limits = TRUE
  ) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  theme_void() +
  ggtitle("OLS Residuals Gradient")
#
ggplot() +
  geom_sf(data = zona_estudio, fill = NA, color = "black") +
  geom_sf(data = data_ols, aes(fill = sem_resid), alpha = 0.8) +
  scale_fill_steps(
    name = "SEM Residuals", low = "lightsteelblue1", high = "tomato1",
    n.breaks = 5, show.limits = TRUE
  ) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  theme_void() +
  ggtitle("SEM Residuals Gradient")
```

## 11. IDW Interpolation of OLS Residuals and Classified Raster
```{r message=FALSE, warning=FALSE}
if (!"fid_1" %in% names(sitios_sem)) {
  data_ols$Residual <- data_ols$ols_resid
} else {
  data_ols$Residual <- data_ols$ols_resid
  join_df <- data_ols %>% st_drop_geometry() %>% dplyr::select(fid_1, Residual)
  sitios_sem <- sitios_sem %>% left_join(join_df, by = "fid_1")
}
#
if (!"Residual" %in% names(sitios_sem)) {
  stop("Column 'Residual' not found in sitios_sem.")
}
#
sitios_centroids <- st_centroid(sitios_sem) %>% dplyr::select(fid_1, Residual, geometry)
centroids_sp <- as(sitios_centroids, "Spatial")
ext <- st_bbox(zona_estudio)
res_x <- 50
raster_base <- raster(xmn = ext["xmin"], xmx = ext["xmax"],
                      ymn = ext["ymin"], ymx = ext["ymax"],
                      crs = st_crs(zona_estudio)$proj4string,
                      resolution = c(res_x, res_x))
idw_model <- gstat::gstat(formula = Residual ~ 1,
                          data = centroids_sp,
                          nmax = 12,
                          set = list(idp = 2))
#
idw_result <- predict(idw_model, as(raster_base, "SpatialGrid"))
raster_resid <- raster(idw_result, layer = "var1.pred")
#
print(raster_resid)
#
vals <- values(raster_resid)
r_mean <- mean(vals, na.rm = TRUE)
r_sd <- sd(vals, na.rm = TRUE)
cat("IDW Raster Mean:", r_mean, " SD:", r_sd, "\n")
classify_sigma <- function(x, m, s) {
  z <- (x - m) / s
  dplyr::case_when(
    z < -2 ~ -2.5,
    z < -1 ~ -1.5,
    z <  1 ~ 0,
    z <  2 ~ 1.5,
    TRUE   ~ 2.5
  )
}
raster_cat <- calc(raster_resid, fun = function(x) {
  classify_sigma(x, r_mean, r_sd)
})
#
sigma_breaks2  <- c(-3, -2, -1, 1, 2, 3)
sigma_labels2  <- c("- >2 σ", "- 1–2 σ", "± 1 σ", "+ 1–2 σ", "+ >2 σ")
sigma_palette2 <- c("#1F5A97", "#82C0D1", "#FFFFFF", "#FFE66D", "#F57C35")
#
map_raster <- tm_shape(raster_cat) +
  tm_raster(
    palette = sigma_palette2,
    breaks  = sigma_breaks2,
    style   = "fixed",
    labels  = sigma_labels2,
    title   = "OLS Residuals (IDW ±σ)"
  ) +
  tm_shape(zona_estudio) +
  tm_borders(col = "black", lwd = 2) +
  tm_layout(legend.outside = TRUE) +
  tm_compass(position = c("RIGHT", "top")) +
  tm_scale_bar(position = c("RIGHT", "bottom"))
#
print(map_raster)
```